---
title: "\\vspace{-0.75in} Final Project: Replicating Tail Risk of Contagious Disease article by Cirillo and Taleb with bootstrapping, frequentist and bayesian inference"
author: \textcolor{Sepia}{Srihari Jaganathan}
date: "STAT 508, Fall 2021"
output: 
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    extra_dependencies: ["flafter"]
geometry: margin=0.75in
header-includes:
 \usepackage{multirow}
 \usepackage{float}
 \usepackage{bbding}
 \usepackage{relsize}
 \usepackage{setspace}
 \usepackage{booktabs}
 \usepackage{siunitx}
 \usepackage[many]{tcolorbox}
 \usepackage{caption}
 \usepackage{hyperref}
 \usepackage{tikz}
 \usepackage{xcolor}
 \usepackage{sectsty}
 \usepackage{adjustbox}
 \usepackage{graphicx}
 \usepackage{amsmath}
 \usepackage{amsfonts}
 \definecolor{brightmaroon}{RGB}{180, 61, 57}
 \definecolor{cornsilk}{RGB}{255, 252, 224}
 \definecolor{aureolin}{RGB}{255, 244, 71}
 \allsectionsfont{\color{brightmaroon}}
 \renewcommand{\topfraction}{.85}
 \renewcommand{\bottomfraction}{.7}
 \renewcommand{\textfraction}{.15}
 \renewcommand{\floatpagefraction}{.66}
 \setcounter{topnumber}{3}
 \setcounter{bottomnumber}{3}
 \setcounter{totalnumber}{4}
 %\floatplacement{figure}{t}
 %\pgfplotsset{compat=1.16}
 
link-citations: yes
bibliography: [packages.bib,references.bib]
linkcolor: blue

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE,tidy.opts=list(width.cutoff=50),tidy=TRUE, message = FALSE,options(scipen=999,digits = 4),dev="pdf")

setwd("/mnt/home/user/stat508/final_project/")

## Libraries

library(cowplot)
library(ggplot2)
library(ggthemes)
library(psych)
library(xlsx)
library(knitr)
library(kableExtra)
library(tibble)
library(tidyverse)
library(evir)
library(eva)
library(scales)
library(reshape2)
library(ggrepel)
library(patchwork)
library(plot3D)
library(MASS)
library(VGAM)
library(formatR)
```

# Introduction

In this study we replicate the article "Tail Risk of Contagious Disease"(@cirillo2020tail). Cirillo and Taleb use probabilistic approach and extreme value theory to show that pandemic deaths are "fat tailed". This is very important work, in risk domain, if a phenomena that we are studying is determined to be fat tailed then its most likely going to be \textbf{unpredictable} and will have destructive consequences. It is very important to note that they published this article in [April 2020](https://arxiv.org/abs/2004.08658) before COVID-19 pandemic started to become widespread. Several researchers have realized this unpredictability phenomena unfortunately very late. For instance below is the quote from Carnegie Mellon University's Delphi Research group which serves as the basis of the CDC’s official communications on COVID-19 forecasting and the blog post was recently published in International Institute of Forecasters(@delphi):

\begin{quote}
\textbf{``}... forecasts of cases and hospitalizations showed repeated, \textbf{sustained lapses in accuracy} for longer-term forecasts, especially at key points during some the larger pandemic waves. Therefore, starting in September 2021, the Hub decided to \textbf{suspend} the inclusion of 2 through 4 week ahead case forecasts and 15 through 28 day ahead hospitalization forecasts in the official ensemble that is generated every week. Modelers and forecasters should continue to innovate and investigate so we can continue to build our understanding of how models can be used to anticipate changes in COVID trends and serve the needs of decision-makers and the general public.\textbf{"}
\end{quote}

\begin{figure}
    \centering
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=\linewidth,height=5cm]{../final_project/images/Pasquale_Cerillo.jpg} % first figure itself
        \caption*{Pasquale Cirillo Source:\href{https://online-learning.tudelft.nl/instructors/pasquale-cirillo/}{Website}}
    \end{minipage}\hfill
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=\linewidth,height=5cm]{../final_project/images/Taleb_mug.jpg} % second figure itself
        \caption*{Nassim Taleb Source:\href{https://en.wikipedia.org/wiki/Nassim_Nicholas_Taleb}{Wikipedia}}
    \end{minipage}
\end{figure}

\begin{tcolorbox}[colback=cornsilk,colframe=aureolin]
Science is about understanding properties, not forecasting single outcomes. - Nassim Taleb
\end{tcolorbox}


As illustrated in Figure \ref{fig:fig0}, most consequential events are not seen in bulk of probability distribution, instead it resides in the tail of the distribution (@taleb2020single). Therefore its important to study the tails. In this study we use tools developed by  @cirillo2020tail and replicate their article. We will develop codes from scratch and not rely on standard packages including, graphical tools to determine if the data generating process is from fat-tailed  distribution, maximum likelihood estimation, estimating uncertainty such as profile likelihood. We will employ techniques learnt in this class such as bootstrapping to estimate parameter uncertainty. The organization of this study is as follows: in the next section, we will provide brief overview of the data, followed by exploratory data analysis. Then we discuss @cirillo2020tail idea of dual distribution which makes it possible to calculate moments and therefore expected value. Next we will employ maximum likelihood estimation and visual approaches to assess the fit of the distribution. To assess the uncertainty of parameter estimates, we will use parametric, non-parametric and bayesian approaches. Finally we conclude the article with implications.



```{r fig0, fig.align = "center",echo=F,dev='tikz',echo=FALSE,fig.cap="\\label{fig:fig0} Probability Density function with unseen rare events beyond threshold $u$",out.width="45%"}
x <- seq(1,10,0.01)
y <- dpareto(x,1,0.1)
df_p <- data.frame(x,y)

shade <- data.frame(x=c(x[x>=6], max(x), 6), y=c(y[x>=6], 0, 0))

ggplot(data=df_p,aes(x=x,y=y)) + theme_bw()+
  geom_line(size=0.6,color="grey39")+ 
  geom_polygon(data = shade, aes(x, y),col="grey39",fill="#FFF8DC") + 
  theme(plot.background = element_blank(),axis.text.x=element_blank(),
        axis.text.y=element_blank(),axis.title=element_text(size=16),
              panel.grid.major = element_blank(),aspect.ratio = 0.7,
              panel.grid.minor = element_blank())+xlab("$x$") + 
  ylab("PDF($x$)")+annotate("segment", x = 7.5, xend = 6.5, y = 0.030, 
                            yend = 0.013, col="grey39",
                            arrow=arrow(length=unit(2.5, "mm")), lwd=0.8)+
   geom_vline(xintercept =6,linetype="dashed",color="blue")+
  annotate(geom="text",x = 4.5, y = 0.035, 
           label = "Unseen Rare Events (Fat Tail)",hjust = 0,size = 6)+
  annotate("text", x = 5.7, y = 0.05, label = "Threshold ($u$)",
           col="blue",hjust = 0,size = 5,angle = 90)
```


# Data 

@cirillo2020tail compiled a list of 72 pandemics with greater than 1000 casualties from 3 different sources (@web1, @wiki, @list). The pandemic dataset (see Table 6 in Appendix)has name of the pandemic, start and end year, average estimated death, upper and lower estimates, rescaled death to today's population, and population at the time of pandemic. We will use average estimated death for the analysis. Table \ref{table:descstat} shows the descriptive statistics of the average estimated deaths. As we can observe the mean and median are farther apart, in addition kurtosis is extremely high indicating that this is a skewed dataset. Table \ref{table:quant} shows a very large difference between $75^{th}$ percentile and the $100^{th}$ percentile indicating heavy skewness. Figure \ref{fig:fig1} shows time series of average estimated deaths in log scale and the histogram of pandemic casualties. One could observe from this figure that most of the pandemics happen after year 1500 and in addition, histogram indicates that there is extreme skewness in the data. In the rescaled death data, some pandemics scaled to current population such as Plague of Justinian and Black Death have over 2 billion casualties which is approximately 30% of current world population.

\begin{table}[h]
\caption{ Descriptive statistics of average estimated deaths ($\times10^3$) in 72 Pandemics}
\centering
\begin{tabular}{lllllllllll}
\hline
                        & n  & Mean & SD    & Median & Trimmed & Min & Max    & Range  & Skew & Kurtosis \\ \hline
Average Estimated Death & 72 & 4878 & 19132 & 82     & 459.9   & 1   & 137500 & 137499 & 5.34 & 31.11    \\ \hline
\end{tabular}
\label{table:descstat}
\end{table}

\begin{table}[h]
\caption{ Quantile of average estimated deaths ($\times10^3$) in 72 Pandemics}
\centering
\begin{tabular}{llllll}
\hline
                         & 0\% & 25\% & 50\% & 75\% & 100\%  \\ \hline
Average Estimated Deaths & 1 & 10   & 82   & 850  & 137500 \\ \hline
\end{tabular}
\label{table:quant}
\end{table}

```{r fig1,fig.align = "center",echo=F,dev='tikz',echo=FALSE,fig.cap="\\label{fig:fig1} Figure (a) shows time series plot of pandemics average estimated deaths in log scale by starting year. Figure (b) shows the histogram of 72 pandemics average estimated deaths which clealy shows heavy skewness in the dataset"}

## Get Data

import_data <- read.xlsx(file = "/mnt/home/user/stat508/final_project/data/long_fat_tail.xlsx",1)
dat <- import_data$avg.est



## Maximum Sum Plot
cum_sum = cumsum(import_data$avg.est)

x_1 <- data.frame(start.year = import_data$start.year,
                  avg.est = import_data$avg.est,cum_events = 1:length(dat))

p1 <- ggplot(x_1, aes(x=start.year, y=avg.est)) +
        geom_point(shape = 4,size = 1.5,col="blue") +
        geom_segment(aes(x=start.year,
                         xend=start.year, y=1, 
                         yend=avg.est),colour = "grey39") + 
        scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
                      labels = trans_format("log10", math_format(10^.x)))+
  annotation_logticks(sides = "l",colour =      "grey39")+ 
        ylab("Average Estimated Deaths ($\\times10^3$)")+
        xlab("Year")+theme_bw() + theme(plot.background = element_blank(),
          panel.grid.major = element_blank(),aspect.ratio = 0.7,
          panel.grid.minor = element_blank(),
          axis.title = element_text(size=rel(0.85))) + 
  ggtitle("(a) Time series plot of pandemics")+
  theme(plot.title = element_text(size = 10),plot.title.position = "plot")


p2 <- ggplot(x_1, aes(x=start.year, y=cum_events)) +
            geom_line(col="grey39") + geom_point(shape = 4,size = 1.5,col="blue") + 
            ylab("Cumulative Number of Pandemics")+
            scale_y_continuous(breaks=seq(0, 72, 10)) +
            xlab("Year")+theme_bw() + theme(plot.background = element_blank(),
    panel.grid.major = element_blank(),aspect.ratio = 0.7,
    panel.grid.minor = element_blank(),
    axis.title = element_text(size=rel(0.85))) +
  theme(plot.title = element_text(size = 10),plot.title.position = "plot")


p3 <- ggplot(data=x_1, aes(x=avg.est/10^5)) + 
             geom_histogram(bins = 72,
                            col="grey39", 
                            fill="grey39", 
                            alpha = .1) + 
              labs( x="Casualties ($\\times10^8$)", y="Count") +
              scale_x_continuous(breaks=seq(0.0, 1.4, 0.2))+
              scale_y_continuous(breaks=seq(0.0, 72, 10))+
              theme_bw() + theme(plot.background = element_blank(),
              panel.grid.major = element_blank(),aspect.ratio = 0.7,
              panel.grid.minor = element_blank(),axis.title = 
                element_text(size=rel(0.85))) + 
  ggtitle("(b) Histogram of casualties") +
  theme(plot.title = element_text(size = 10),plot.title.position = "plot")

plot_grid(p1,p3,nrow = 1,rel_widths = c(1,1),align = "vh")

```

# Exploratory Data Analysis

When we suspect heavy skewness and also large outlying observations, one needs to further conduct expolratory analysis to determine of the underlying data generating process is from heavy tailed distribution such as Pareto distribution.Cirillo (@cirillo2013your) provides an excellent overview and diagnostics to determine of the distribution has a heavy tailed phenomena and rule out other confounding distributions such as Log-normal distribution. In this article we will use diagnostic plots as proposed by @cirillo2020tail. If the data has a heavy tail, one would observe a convex shape in an exponential Q-Q plot.  Figure \ref{fig:fig2} (a) shows the Exponential Q-Q plot against the observed casualties, as we could clearly see that there is a convex shape which may suggest presence of heavy tail. The second useful plot is the maximum to sum ratio plot. For a random variable that is identical and independently distributed (i.i.d), according to law of large numbers if $E[X^p] < \infty$ the ratio ${M_n^p}/{S_n^p}$ would converge to zero for order $p=1,2,3,...$ as $n \rightarrow \infty$. Here $p$ is the moments, $M_n = max(X_1^p,X_2^p,...,X_n^p)$ is partial maximum value at $n$ for $p$ similarly $S_n^p = \sum_{i=1}^{n} X_i^p$.

\begin{tcolorbox}[colback=cornsilk,colframe=aureolin]
Exponential Q-Q, maximum to sum ratio, survival and mean excess (residual life) plots are useful diagnostic tools to determine if the data is from a heavy tail distribution.
\end{tcolorbox}


```{r fig2,fig.align = "center",echo=F,dev='tikz',echo=FALSE,fig.cap="\\label{fig:fig2} (a) shows Exponential Q-Q plot against observed data. Convex shape indicates heavy tail. (b) shows Maximum to Sum plot, and all the moments do not converge to 0 indicating strong presence of heavy tail"}

exp_qq <- data.frame(x=sort(dat)/10^5,y=qexp(ppoints(dat)))
mod <- lsfit(sort(dat/10^5),qexp(ppoints(dat)))

p4 <- ggplot(exp_qq, aes(x=x, y=y)) +
        geom_point(shape = 1,size = 2,col="grey39") + scale_x_continuous(breaks=seq(0.0, 1.4, 0.2)) +
        labs( x= "Casualties ($\\times10^8$)", y="Exponential Quantiles") +
              theme_bw() + theme(plot.background = element_blank(),
              panel.grid.major = element_blank(),aspect.ratio = 0.7,
              panel.grid.minor = element_blank())+
  geom_abline(intercept = mod$coefficients[1],
              slope= mod$coefficients[2],col="grey39")+ ggtitle("(a)")+
  theme(plot.title = element_text(size = 10)) + 
  ggtitle("(a) Exponential Q-Q plot") + 
  theme(plot.title = element_text(size = 10),plot.title.position = "plot")

msrd <- data.frame( x = seq_along(dat),
                    p1 = (cummax(dat^1)/cumsum(dat^1)),
                    p2 = (cummax(dat^2)/cumsum(dat^2)),
                    p3 = (cummax(dat^3)/cumsum(dat^3)),
                    p4 = (cummax(dat^4)/cumsum(dat^4)))
msrd.gg <- melt(msrd ,  id.vars = 'x', variable.name = 'series')
msrd.gg$series <- factor(msrd.gg$series, levels=c("p1", "p2", "p3","p4"), 
                         labels=c("p=1", "p=2", "p=3","p=4"))

p5 <-   msrd.gg %>%
  mutate(label = if_else(x == max(x), as.character(series), NA_character_))  %>%
  ggplot(aes(x = x, y = value, group = series, colour = series)) + 
  geom_line() + scale_x_continuous(limits=c(0,85),breaks=seq(0.0, 100, 10))+
  scale_y_continuous(limits=c(0,1),breaks=seq(0.0, 1, 0.1))+
  xlab("Observations (n)")+
  ylab('$\\frac{M_n^p}{S_n^p}$')+ #expression(paste(frac(M[n]^p, S[n]^p)))
  geom_label_repel(aes(label = label),size = 3,
                  nudge_x = 1,box.padding = 0,label.size=NA,fill = NA,
                  na.rm = TRUE) +theme_bw() + theme(plot.background = element_blank(),
              panel.grid.major = element_blank(),aspect.ratio = 0.7,
              panel.grid.minor = element_blank(),legend.position = "none",
              axis.title.y = element_text(angle = 0,vjust = 0.5))+ 
  ggtitle("(b)") + theme(plot.title = element_text(size = 10))+ 
  ggtitle("(b) Maximum to Sum ratio plot") +
  theme(plot.title = element_text(size = 10),plot.title.position = "plot")


plot_grid(p4,p5,nrow = 1,align = "vh")
```

Log-Log survival plot or Zipf plot is another useful diagnostic tool to determine if the data is heavy tailed. Figure \ref{fig:fig3} a.  The Zipf plot shows a clear linear downward trend, indicating the presense of heavy tailed distribution. Non parametric hill estimator plot can be used to estimate the tail parameter $\hat{\xi}_n$ which will be discussed in subsequent sections. Let $X_1,X_2,X_3,...,X_n$ be i.i.d random variable, and the corresponding order statistics be $X_{n,n} \le ... \le X_{1,n}$, then the tail parameter $\hat{\xi}_n$ can be estimated as follows: 

\begin{equation}
\hat{\xi}_n = \frac{1}{k} \sum_{i=1}^{k} log(X_{i,n}) - log(X_{k,n}), 2 \le k \le n.
\end{equation}

```{r fig3,fig.align = "center",echo=F,dev='tikz',echo=FALSE,fig.cap="\\label{fig:fig3} (a) shows Log-Log survival or Zipf plot. A linear trend down indicates a heavy tailed distribution. (b) shows the hill plot, a value of greater than 1 indicates a heavy tailed distribution."}

ldat = log(sort(dat,decreasing = T))
n <- length(dat)
ns = 1:(n - 1)
alpha = 0.05

xi = cumsum(ldat[-n])/ns - ldat[-1]

xi_se = xi/sqrt(ns)

lci <- xi - qnorm(1 - alpha/2) * xi_se
hci <- xi + qnorm(1 - alpha/2) * xi_se

hill_est <- data.frame(ns,xi,lci,hci)

hill_est_df <- melt(hill_est,id.vars = 'ns', variable.name = 'series')


label <- sort(dat,decreasing = T)[c(1, seq(10,70,10))]

p6 <- ggplot(hill_est_df, aes(x=ns, y=value, group=series)) +
  geom_line(aes(linetype=series,colour = series))+ 
  scale_linetype_manual(values=c("solid","dashed", "dashed"),labels = c('xi' = '$\\xi$','lci'= "LCL",'hci'="HCL"))+ 
  theme_bw() + scale_color_manual(values=c("grey39", "grey39", "grey39")) + theme(plot.background = element_blank(),
          panel.grid.major = element_blank(),legend.position="none",aspect.ratio = 0.7,
          panel.grid.minor = element_blank()) + 
  scale_x_continuous(limits=c(1,71),breaks=c(1, seq(10,70,10)),sec.axis = dup_axis( name = "Threshold",breaks=c(1, seq(10,70,10)),labels = label),guide = guide_axis(check.overlap = TRUE))+
  scale_y_continuous(limits=c(-1,6),breaks=seq(-1, 6, 1)) +xlab("Ordered Observations") +ylab('$\\xi$ with 95$\\%$ CI') +guides(colour="none") + theme(legend.title=element_blank()) + ggtitle("(b) Hill Estimator") +theme(plot.title = element_text(size = 10),plot.title.position = "plot")


s_eq <- 1-ppoints(dat)

data_final <- data.frame(x = sort(dat),y=s_eq)

mod1 <- lsfit(s_eq,sort(dat))

mod1  <- lm(log10(s_eq) ~ log10((sort(dat))))

  p7 <- ggplot() + 
  geom_point(data = data_final, aes(x = x, y = y), color = "grey39",shape=21,size=2,fill="grey70") +
  scale_x_log10(
    breaks = scales::trans_breaks("log10", function(x) 10^x),
    labels = scales::trans_format("log10", scales::math_format(10^.x))
  ) + scale_y_log10(breaks = c(1,0.5,0.1,0.05,0.01,0), limits = c(0.0069, 1))+annotation_logticks(sides = "trbl",colour = "grey39")+
  theme_bw() + theme(plot.background = element_blank(),
              panel.grid.major = element_blank(),aspect.ratio = 0.7,
              panel.grid.minor = element_blank()) +
  xlab("Casualties ($\\times 10^3$)") +
  ylab("S($x$)") +geom_abline(intercept = mod1$coefficients[1],slope= mod1$coefficients[2],col="grey39") + ggtitle("(a) Log-Log Survival or Zipf plot") +theme(plot.title = element_text(size = 10),plot.title.position = "plot")

plot_grid(p7,p6,nrow = 1,rel_widths = c(1,1),align = "vh")
```

Mean excess plot or mean residual life plot can be used to determine presence of fat tail in addition to also determining threshold value $u$. An upward linear trend indicates presence of fat tail. Mean excess plot is calculated using following equation.

\begin{equation}
\hat{M}(u) = \frac{\sum_{i=1}^{n} X_{i} \mathbb{I}_{[X_i>u]}} {\sum_{i=1}^{n}\mathbb{I}_{[X_i>u]} }
\end{equation}


Figure \ref{fig:fig3a} a shows that there is a strong upward trend. Whenever there is change in direction or plateauing that may be use to select threshold. In this instance @cirillo2020tail choose the threshold of 200 as on optimal threshold.   Figure \ref{fig:fig3a} b shows data that is above the threshold of 200. Table \ref{table:tab_char} provides a summary of all diagnostic tools used to identify fat-taildness in the data generating process.


```{r fig3a,fig.align = "center",echo=F,dev='tikz',echo=FALSE,fig.cap="\\label{fig:fig3a} (a) shows the mean-excess plot at various threshold $u$. An upward linear trend indicates presence of fat tail. (b) shows time series plot with choosen observation above the threshold 200($\\times 10^3$) casualties"}

meplot=function(data,cut=5) {
      # In cut you can specify the number of maxima you want to exclude.
      # The standard value is 5
      data=sort(as.numeric(data));
      n=length(data);
      mex=c();
      
      for (i in 1:n) {
        mex[i]=mean(data[data>data[i]])-data[i];
      }
      data_out=data[1:(n-cut)];
      mex_out=mex[1:(n-cut)];
      
      return(cbind(data_out,mex_out))
      
}

me_data <- data.frame(meplot(dat))

p8 <- ggplot() + 
          geom_point(data = me_data, aes(x = data_out, y = mex_out), color = "grey39",shape=21,size=2,fill="grey70") +theme_bw() + theme(plot.background = element_blank(),
              panel.grid.major = element_blank(),aspect.ratio = 0.7,
              panel.grid.minor = element_blank()) + xlab("Threshold $u$")+
  ylab("Mean Excess E($u$)") + ggtitle("(a) Mean Exess plot") +theme(plot.title = element_text(size = 10),plot.title.position = "plot")
  

p9 <- ggplot(x_1,aes(x=start.year, y=avg.est)) + 
        geom_segment( aes(x=start.year, xend=start.year, y=1, yend=avg.est),colour = "grey39") + geom_point(data = x_1[x_1$avg.est>=200,], shape = 21,size = 1.5,col="blue",fill="skyblue") +
        scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
                      labels = trans_format("log10", math_format(10^.x)))+annotation_logticks(sides = "l",colour =      "grey39")+ geom_hline(yintercept=200,col="grey39",linetype = "dashed")+
        ylab("Average Estimated Deaths ($\\times10^3$)")+ annotate("text", x=-400, y=300, label= "$u$ = 200",hjust = 0,size = 3) +
        xlab("Year")+theme_bw() + theme(plot.background = element_blank(),
          panel.grid.major = element_blank(),aspect.ratio = 0.7,
          panel.grid.minor = element_blank(),axis.title = element_text(size=rel(0.85))) +  ggtitle("(b) Time series with threshold") +theme(plot.title = element_text(size = 10),plot.title.position = "plot")

plot_grid(p8,p9,nrow = 1,rel_widths = c(1,1),align = "vh")
```

\begin{table}[h!]
\caption{Charectersitic of fat tail of various diagnostic plots}
\centering
\begin{tabular}{ll}
\hline
Diagnostic Plot                 & Characteristic of Fat Tail           \\ \hline
Maximum to Sum Ratio Plot       & Non-convergence to zero of moments   \\
Exponential Q-Q Plot            & Convex Shape                         \\
Log-Log Survival plot/Zipf Plot & Linear downward trend                \\
Mean Excess Plot                & Upward linear trend                  \\
Hill Estimator of $\xi$         & Estimator $\xi$ \textgreater 1 \\ \hline
\end{tabular}
\label{table:tab_char}
\end{table}


# Dual Data

Lack of moments as shown in the maximum to sum plot in Figure \ref{fig:fig2} (b)  does not mean that we have infinite mean, as the world is bounded by the population. @cirillo2016statistical and @cirillo2016expected proposed dual data approach with special log transformation. Let L be the lower bound of pandemic fatalities and H be the maximum possible fatalities which could be the world population, then the dual data is obtained by following transformation. 

\begin{equation}
\varphi(Y) = L - H \ log\left(\frac{H-Y}{H-L} \right )
\end{equation}

from which $\varphi(Y) \in C^\infty$, $\varphi^{-1}(Y) =H$ and $\varphi^{-1}(L) = \varphi(L) = L$. Figure \ref{fig:figpdf1} shows what would happen if ignores the upper existence of upper bound H, since only M is observed. The new random variable is defined as $Z = \varphi(Y)$ with lower bound $L$ and infinite upper bound. The expected value for random variable $Y$ can be obtained as follows:

\begin{equation}
E[Y] = (H-L) e^{\frac{\frac{1}{\xi}\sigma}{H}}\left(\frac{\sigma}{H\sigma}\right)^{\frac{1}{\xi}}\Gamma\left(1-\frac{1}{\xi},\frac{\sigma}{H\xi}\right)+L
\end{equation}

where $\sigma = \beta(n_u/n)^\xi$, $n_u$ is the number of opeservations abover threshold and $n$ is the total number of observations. Parameters $\beta$ and $\xi$ can be estimated using maximum likelihood estimation.


\begin{figure}[H]
    \centering
    \includegraphics[height=6cm]{../final_project/ggplot_mod-crop.pdf}
    \caption{Figure showing a Log-Log plot of what would happen if one ignores the upper bound H, since only M is observed}
    \label{fig:figpdf1}
\end{figure}

\begin{tcolorbox}[colback=cornsilk,colframe=aureolin]
Dual data approach via special log transformation helps us calculate 'shadow' moments such as expected pandemic fatalaties which is not possible when using raw data.
\end{tcolorbox}


# Modeling with Extreme Value Theory using threshold models

As @cirillo2020tail mention in the article, most important information is in tail of the distribution. They employ tools from extreme value theory to study pandemic casualties. Generalized Pareto distribution was used to fit the data for dual transformed random variable $Z$. The cumulative distribution function $F(z)$ and the probability density function $f(z)$ is given by following equation. 

\begin{equation}
F(z) = \begin{cases}
1 - \left(1+ \frac{\xi(z-u)}{\beta}\right)^{-1/\xi} & \text{for }\xi \neq 0, \\
1 - e^{ \left(-\frac{z-u}{\beta}\right)} & \text{for }\xi = 0,
\end{cases} \\
\end{equation}

\begin{equation}
f(z) = \frac{1}{\beta}\left(1 + \frac{\xi (z-u)}{\beta}\right)^{\left(-\frac{1}{\xi} - 1\right)}
\end{equation}


@cirillo2016statistical recommend using $u = 200,000$ victims as appropriate threshold. This leads us to $nu =25$ (34.7 $\%$)observations that are above 200,000 observations. The key parameter of interest is the shape parameter $\xi$. As noted in previous sections, if $\xi>1$ then there is no finite moments such as mean. With dual transformations we are able to estimate the mean since random variable $Z$ has finite support between $L$ and $H$. As far as threshold $u$,one could use graphical tools to determine appropriate values. As outlined in the exploratory data analysis, one could rely on graphical tools such as Zipf plot and mean excess plot.   

## Fitting the data with Maximum likelihood

Maximum Likelihood estimation (@coles2001introduction) methods are appropriate for cases where $\xi > 0$. Excellent overview of various estimation methods to fit generalized Pareto distribution are provided in de2010parameter1 and de2010parameter2. Following is the log likelihood equation to estimate parameters $\theta = (\xi,\beta)$.  

\begin{equation}
\mathcal{L}_n = \sum_{i=1}^{n} \text{log } f(X_i;\xi,\beta)
\end{equation}


where $\xi$ and $\beta$ can be estimated by ($\hat{\xi},\hat{\beta}$) = $\text{argmax}_{\xi,\beta} \mathcal{L}_n$. After applying numerical optimization to maximize the log likelihood, we obtain the estimates and standard errors (SE) as outlined in Table \ref{table:tab_mle}. The estimates are identical to the one obtained in the @cirillo2016statistical. 3D surface and contour plot is shown in \ref{fig:figpdf2}. 3D surface area around the optimum estimates are very flat and not steep which indicates that there will be high level of uncertainty and its reflected in large standard errors. 

\begin{table}[h!]
\caption{Comparison of estimates from this analysis and Cirillo and Taleb (2020)}
\centering
\begin{tabular}{lllllll}
\hline
\multirow{2}{*}{}   &  & \multicolumn{2}{l}{This Analysis} &  & \multicolumn{2}{l}{Cirillo and Taleb (2020)} \\ \cline{3-4} \cline{6-7} 
                    &  & Estimate         & SE             &  & Estimate               & SE                  \\ \cline{1-1} \cline{3-4} \cline{6-7} 
$\beta \times 10^3$ &  & 1174.72          & 535.08         &  & 1174.7                 & 536.5               \\
$\xi$               &  & 1.62             & 0.52           &  & 1.62                   & 0.52                \\ \hline
\end{tabular}
\label{table:tab_mle}
\end{table}

\begin{figure}[H]
    \centering
    \begin{minipage}{0.45\textwidth}
    \centering
      \includegraphics[width=3.35in,height=8cm]{../final_project/gpd1_tikz-crop.pdf}
     \end{minipage} 
    \begin{minipage}{0.45\textwidth}
    \centering
      \includegraphics[width=3.35in,height=8cm]{../final_project/gpd_contour_tikz-crop.pdf}
     \end{minipage}
     \caption{Left figure shows the 3D surface plot of log likelihood function with contours, as can be seen the surface plot is flat and not steep which will indicate there will be high level of uncertainity in estimates. Right figure shows the contour plot with optimum values highlighted at $\beta=1174.72$ and $\xi=1.62$.}
    \label{fig:figpdf2}
\end{figure}


## Assesing the fit of the distribution

There are two approaches to assess the fit of generalized Pareto distribution (GPD) to empirical data. First one is visual inspection of fit vs actual data and the second is goodness of fit test (@choulakian2001goodness). In this analysis we use visual inspection to assess the fit of the distribution. Figure \ref{fig:est} (a) shows the Log-Log survival plot of fitted vs actual data. The fit appears to be very good. Similarly fit is realized when we use CDF of the fitted vs actual data as shown in Figure \ref{fig:est} (b).   

```{r est,fig.align = "center",echo=F,dev='tikz',fig.height= 3.5,echo=FALSE,fig.cap="\\label{fig:est} (a) shows the mean-excess plot at various threshold $u$. An upward linear trend indicates presence of fat tail. (b) shows time series plot with choosen observation above the threshold 200($\\times 10^3$) casualties"}

### Dual Data ###

L <- 1
H <- 7700000
dual = L -  H* log( (H-dat) / (H-L) )

### Log Likelihood of GPD ####
llfun <- function(parms,threshold=200,indata=dual){
  
  scale <- parms[1]
  shape <- parms[2]
  loc <- 0
  
  x <- indata[indata>= threshold] - threshold
  
  llik <- log(1/scale * (1 + shape * (x - loc)/scale)^(-1/shape - 1))
  
  return(-sum(llik))
  }

### optimize LLik ###
opt.gpd <- optim(fn=llfun,par = c(1000,1),lower = c(0.001,0.001),method="L-BFGS-B",threshold=200,indata=dual,hessian = T)

### Extract Paramaeters and calculate SE ###
opt.par <- opt.gpd$par
vcovx <- solve(opt.gpd$hessian)
opt.se <- sqrt(diag(vcovx))

beta <- opt.par[1]
xi <- opt.par[2]

### Fit of distribution ###


## Actual Vs Theoretical Plot ##
x_200 = sort(dual[dual>=200])

qf <- evir::qgpd(ppoints(x_200), xi = xi,beta = beta,mu = 200)


p10 <- ggplot2::qplot(x_200,qf) + geom_point(aes(x = x_200, y = qf), color = "grey39",shape=21,size=2,fill="grey70") + scale_x_log10(breaks = scales::trans_breaks("log10", function(x) 10^x),
    labels = scales::trans_format("log10", scales::math_format(10^.x))) + scale_y_log10(
    breaks = scales::trans_breaks("log10", function(x) 10^x),labels = scales::trans_format("log10", scales::math_format(10^.x)))+annotation_logticks(sides = "trbl",colour = "grey39")+ theme_bw() + geom_smooth(method="lm",se=FALSE,col="blue",size=0.4) + theme(plot.background = element_blank(),
              panel.grid.major = element_blank(),aspect.ratio = 0.7,
              panel.grid.minor = element_blank()) + xlab("Actual Data") + ylab("Fitted Data")+  ggtitle("(a)") +theme(plot.title = element_text(size = 10),plot.title.position = "plot")

### Survival Plot ###

surv_emp <- 1-ppoints(x_200)
surv_fit <- 1-evir::pgpd(x_200, xi = xi,beta = beta,mu = 200)

p11 <- ggplot2::qplot(x_200,surv_emp) + geom_point(aes(x = x_200, y = surv_emp), color = "grey39",shape=21,size=2,fill="grey70") + scale_x_log10() + scale_y_log10(breaks = c(1,0.5,0.1,0.05,0.01,0), limits = c(0.0069, 1))+annotation_logticks(sides = "trbl",colour = "grey39")+ theme_bw() + theme(plot.background = element_blank(),panel.grid.major = element_blank(),aspect.ratio = 0.7,
              panel.grid.minor = element_blank()) + xlab("Actual Data") + ylab("Survival S($x$)") +  geom_line(data = data.frame(x=x_200,y=surv_fit),mapping = aes(x = x, y = y),col="blue",size=0.4)+ggtitle("(a) Log-Log survival plot") +theme(plot.title = element_text(size = 10),plot.title.position = "plot")



p12 <- ggplot2::qplot(x_200,(1-surv_emp)) + geom_point(aes(x = x_200, y = (1-surv_emp)), color = "grey39",shape=21,size=2,fill="grey70") + scale_x_log10() +annotation_logticks(sides = "tb",colour = "grey39")+ theme_bw() + theme(plot.background = element_blank(),
              panel.grid.major = element_blank(),aspect.ratio = 0.7,
              panel.grid.minor = element_blank()) + xlab("Actual Data (Log Scale)") + ylab("CDF F($x$)") +  geom_line(data = data.frame(x=x_200,y=(1-surv_fit)),mapping = aes(x = x_200, y = y),col="blue",size=0.4)+ggtitle("(b) CDF of Fit vs. Actual") +theme(plot.title = element_text(size = 10),plot.title.position = "plot")

plot_grid(p11,p12,nrow = 1,rel_widths = c(1,1),align = "vh")


```

\begin{tcolorbox}[colback=cornsilk,colframe=aureolin]
Maximum likelihood estimation is appropriate when shape parameter $\xi>0$. Visual inspection of fitted vs actual and goodness of fit test statistic such as Anderson-Darling or bootstrap test can be used to assess the fit of GPD to the data.
\end{tcolorbox}


# Uncertainity of shape parameter $\xi$

In this section, we study the uncertainty of shape parameter $\xi$. We employ  non-parametric bootstrapping, profile likelihood, parametric bootstrapping, simulation using asymptomatic normality property of maximum likelihood inference, and finally bayesian inference to understand the uncertainty of shape parameter $\xi$. 

## Non-parametric bootstrapping

The pandemic dataset was based on historical reporting which can be inaccurate. There are two forms of inaccuracies that could occur, one could be under/over reporting and the other could be missing information. Recognizing this, @cirillo2016statistical perturbed data by randomly varying the data by $\pm 20\%$ and resampled 10,000 times and estimated parameter $\hat\theta = (\hat\xi,\hat\beta)$. Frequency histogram of this estimate is shown in Figure \ref{fig:nparm} (a). Results are very similar to the one reported in @cirillo2016statistical and its robust data perturbation i.e., the estimates are very close to maximum likelihood estimation. In addition, @cirillo2016statistical jackknifed data by making randomly missing $1\%$ to $10\%$ and estimating the parameters is shown in Figure \ref{fig:nparm} (b). The estimates are closer to the MLE estimates and very the parameter is very robust to missing values.


```{r nparm,fig.align = "center",echo=F,dev='tikz',fig.height= 3.5,echo=FALSE,fig.cap="\\label{fig:nparm} (a) Frequency histogram of estimated $\\xi$ parameter from 10,000 resampled data perturbed by $\\pm 20\\%$ (b)Frequency histogram of jackknifed data by making randomly missing $1\\%$ to $10\\%$. MLE estiamtes are shown in red dashed lines at 1.62"}

## Non Parametric

n <- 10000
tol <- 0.2
len <- length(dual)

## Perturbed simulated data

sim_per <- matrix(NA,n,3)

for (i in 1:n){
  
  runi <- runif(len,-0.2,0.2)
  x_200_dist <- dual + dual*runi
  
  sim_opt <- optim(fn=llfun,par = c(1000,1),threshold=200,
  indata=x_200_dist,hessian = T)
  sim_per[i,1] <- sim_opt$par[1]
  sim_per[i,2] <- sim_opt$par[2]
  sim_per[i,3] <- sim_opt$value
}


df.sim <- data.frame(sim = sim_per[,2])

p13 <- ggplot(df.sim, aes(x=sim)) + 
  geom_histogram(aes(y=..ndensity..), colour="black", fill="#FFF8DC",bins = 20)+
  geom_vline(xintercept = xi,col="red",linetype = "dashed")+theme_bw() + theme(plot.background = element_blank(),
              panel.grid.major = element_blank(),aspect.ratio = 0.7,
              panel.grid.minor = element_blank()) + xlab("$\\xi$") +ylab("Probability")+ggtitle("(a)") +theme(plot.title = element_text(size = 10),plot.title.position = "plot")

#FFF8DC
### Jackknife data ###

sim_jk <- matrix(NA,n,3)

for (i in 1:n){
  sel <- sample(1:72,sample(65:71,1),replace = F)
  x_dist <- dual[sel]
  #yu <- fit.gpd(average_200_dist, threshold = 200, method = 'Grimshaw', show = F)
  yu <-optim(fn=llfun,par = c(1000,1),threshold=200,
  indata=x_dist,hessian = T)
  sim_jk[i,1] <- yu$par[1]
  sim_jk[i,2] <- yu$par[2]
  sim_jk[i,3] <- length(sel)
  }

df.sim.jk <- data.frame(sim=sim_jk[,2])



p14 <- ggplot(df.sim.jk, aes(x=sim)) + 
        geom_histogram(aes(y=..ndensity..), colour="black", fill="#F0F8FF",bins = 20)+
        geom_vline(xintercept = xi,col="red",linetype = "dashed")+theme_bw() + xlim(1.2,2) +
        theme(plot.background =    element_blank(),panel.grid.major = element_blank(),aspect.ratio = 0.7,
              panel.grid.minor = element_blank()) + xlab("$\\xi$") +ylab("Probability")+ggtitle("(b)") +theme(plot.title = element_text(size = 10),plot.title.position = "plot")

plot_grid(p13,p14,nrow = 1,rel_widths = c(1,1),align = "vh")

```

## Profile likelihood

Profile likelihood is an excellent approach to estimate confidence interval of estimates. @Suhasini provides an excellent review of profile likelihood. Wilks theorm gives us the following equation:

\begin{equation}
2 \{ \mathcal{L}_n(\hat\xi_n,\hat\beta_n) - \mathcal{L}_n(\xi_0,\beta_{\xi_0})\} \xrightarrow{D} \chi^2_p 
\end{equation}

where $\xi_0$ is a true parameter, and $\chi^2_p$ is Chi-square distribution with $p$ degrees of freedom representing $\xi_0$ paramater. Using this equation we can construct the confidence interval with $(1-\alpha)$ level of significance and is given by:

\begin{equation}
2 \Big\{ \mathcal{L}_n(\hat\xi_n,\hat\beta_n) - \mathcal{L}_n(\xi,\beta_\xi)\Big\} \le \chi^2_p(1-\alpha)
\end{equation}

Figure \ref{fig:prof} (a) shows the profile log likelihood. Confidence interval at $95\%$ for $\xi$ parameter is (0.84,3.04) highlighted in blue. 

```{r prof,fig.align = "center",echo=F,dev='tikz',fig.height= 3.5,echo=FALSE,fig.cap="\\label{fig:prof} (a) Profile likelihood estimation of $\\xi$ with 95$\\%$ confidence interval (b) Paramteric bootstrapping: $84\\%$ of observations are $\\xi > 1$"}

#insipired by: https://www.r-bloggers.com/2015/11/profile-likelihood/

### Pofile Log Likelihood ###
prof_log_lik=function(a){
   b=(optim(1,function(z) llfun(c(z,a)),method = "Brent",lower = 10,upper=10e6))$par
   return(llfun(c(b,a)))
 }

vx=seq(.01,5,length=101)

vl=-Vectorize(prof_log_lik)(vx)

## Max Like via profile likelihood ##

v1 = optim(1,prof_log_lik)
v_par = v1$par
v_val = -v1$value

## Likelihood ratio at 95% confidence interval
h1 =-optim(1,prof_log_lik)$value-qchisq(.95,1)/2
b1=uniroot(function(z) Vectorize(prof_log_lik)(z)+h1,c(.5,1.5))$root
b2=uniroot(function(z) Vectorize(prof_log_lik)(z)+h1,c(1.6,4))$root

p15 <- ggplot2::qplot(vx,vl,geom = "line") +geom_line(col="grey39") + 
                theme_bw() + theme(plot.background =    element_blank(),
                                   panel.grid.major = element_blank(),aspect.ratio = 0.7,
                                   panel.grid.minor = element_blank()) +xlab("$\\xi$") + 
                                  ylab("Profile Log Likelihood")+ 
                                  geom_vline(xintercept = v_par,col="red",linetype = "dashed")+
  geom_hline(yintercept=v_val,col="grey39",linetype = "dashed") + 
  geom_hline(yintercept = h1,col="grey39",linetype = "dashed") + 
  geom_segment(aes(x = x1, y = y1, xend = x2, yend = y2),
               data = data.frame(x1 =b1, x2 = b2, y1 = h1, y2 = h1),col="blue") + 
  geom_segment(aes(x = x1, y = y1, xend = x2, yend = y2),data = data.frame(x1 =b1, x2 = b2, y1 = h1, y2 = h1),col="blue") + 
  geom_segment(aes(x = x1, y = -Inf, xend = x1, yend = y2),data = data.frame(x1 =b1, x2 = b2, y1 = h1, y2 = h1),col="blue",linetype = "dashed") + 
  geom_segment(aes(x = x2, y = -Inf, xend = x2, yend = y2),data = data.frame(x1 =b1, x2 = b2, y1 = h1, y2 = h1),col="blue",linetype = "dashed") + 
  annotate("text", x = 0.8, y = -265, label = "0.84",col="blue",hjust = 1,size = 3) + 
  geom_segment(aes(x = b1, y = -265, xend = b2, yend = -265), arrow = arrow(length = unit(0.3, "cm"),ends="both"),col="blue") + 
  annotate("text", x = 3.1, y = -265, label = "3.04",col="blue",hjust = 0,size = 3) + annotate("label", x = 1.94, y = -265, label = "95$\\%$ CI",col="blue",size = 3, label.size=NA,fill="white") + scale_y_continuous(limits = c(-270, -240))+ggtitle("(a)") +theme(plot.title = element_text(size = 10),plot.title.position = "plot")



### Parametric Bootstrapping ###

set.seed(1)
nsim = 10000
sim1 <- matrix(NA,nsim,2)

for (i in 1:nsim){
  rg <- evir::rgpd(25, xi = xi,mu = 200,beta = beta)
  gp <- optim(fn=llfun,par = c(1000,1),threshold=200,
  indata=rg,hessian = T)
  
  sim1[i,1] <- gp$par[1]
  sim1[i,2] <- gp$par[2]

  #print(gp$estimate[2])  
  }

df.sim1 <- data.frame(sim1)

p16 <- ggplot(df.sim1, aes(x=X2)) + 
        geom_histogram(aes(y=..ndensity..), colour="black", fill="#DBD7D2",bins = 20)+
        geom_vline(xintercept = xi,col="red",linetype = "dashed",size=0.7)+theme_bw() + 
        theme(plot.background =    element_blank(),panel.grid.major = element_blank(),aspect.ratio = 0.7,panel.grid.minor = element_blank()) + xlab("$\\xi$") +ylab("Probability") + scale_x_continuous(limits = c(0, 5))+ggtitle("(b) Parametric boostraping") +theme(plot.title = element_text(size = 10),plot.title.position = "plot")+ geom_vline(xintercept = 1,linetype = "dashed",col="blue")

plot_grid(p15,p16,nrow = 1,rel_widths = c(1,1),align = "vh")

```

## Parametric Bootstrapping

Parametric bootstrapping is a computationally intensive method, to estimate uncertainity of $\xi$ parameter. The steps involved parametric bootstrapping are as follows:

1. Estimate $\hat\theta = (\hat\xi,\hat\beta)$ using MLE.
2. Based on the MLE estimates $\hat\theta$ , generate random number of sample size $n_u=25$ (number of observations above threshold $u$ from GPD.
3. Apply MLE and store the estimated parameter values as $\hat\theta_i^* = (\hat\xi_i^*,\hat\beta_i^*)$.
4. Repeat steps 2 to 3 for $i=1,2,...,n$ times, here we set $n=10,000$.
5. Calculate the quantity of interest based on step 4. In our instance it is the frequency distribution of $\xi$.

Figure \ref{fig:prof} (b) shows the frequency histogram of paramteric bootstrapping. We observe that approximately $84\%$ of observations are $>1$. 


## Asymptomatic normality

Based on the properties of maximum likelihood estimation we can assume asymptomatic normality of the estimates when the sample size is large. Using this property we simulate using multivariate random normal distribution with mean as $\hat\theta = (\hat\xi,\hat\beta)$ and variance as $\hat{V}(\hat\theta)$ which is the inverse of Fisher information matrix (Hessian at optimum) which is readily available in any optimization routine. The following equation gives us the simulated estimates of parameters $\overset{\sim}{\theta} = (\overset{\sim}{\xi},\overset{\sim}{\beta})$


\begin{equation}
\overset{\sim}{\theta} \sim \mathcal{MVN}(\hat\theta,\hat V(\hat\theta))
\end{equation}


Figure \ref{fig:bayes} (a) shows the frequency histogram of $\hat\xi$. Greater than $95\%$ of values are greater than 1, and almost $>99.99\%$ values are greater than $0$. 

## Bayesian Inference

We use Bayesian estimation to estimate parameter uncertainty. `rstan` package (@rstan) in `R` to estimate posterior density of parameters. @Vehtari has programmed all functions related to GPD in `Stan`. We used this code to model GPD for Bayesian inference. Figure \ref{fig:bayes} (b) shows the posterior of $\hat\xi$. $95\%$ of values are greater than 1, and $100\%$ values are greater than $0$. 


```{r bayes,fig.align = "center",echo=F,dev='tikz',fig.height= 3.5,echo=FALSE,fig.cap="\\label{fig:bayes} (a) Shows the simulated estimates $\\xi$ based on asymptomatic normality and (b) shows the bayesian posterier density"}

load("params.rda")

bayes.parms <- data.frame(xi=params$k)

p17 <- ggplot(bayes.parms, aes(x=xi)) + 
        geom_histogram(aes(y=..ndensity..), colour="black", fill="#D0F0C0",bins = 20)+
        geom_vline(xintercept = xi,col="red",linetype = "dashed")+theme_bw() + 
        theme(plot.background =    element_blank(),panel.grid.major = element_blank(),aspect.ratio = 0.7,
              panel.grid.minor = element_blank()) + xlab("$\\xi$") +ylab("Probability") + scale_x_continuous(limits = c(0, 5))+ggtitle("(b) Bayesian posterier density") +theme(plot.title = element_text(size = 10),plot.title.position = "plot")


mvnorm_freq <- MASS::mvrnorm(n=10000,mu = opt.par,Sigma = vcovx )

df_mvnorm <- data.frame(beta = mvnorm_freq[,1],xi = mvnorm_freq[,2])

p18 <- ggplot(df_mvnorm, aes(x=xi)) + 
        geom_histogram(aes(y=..ndensity..), colour="black", fill="#F4C2C2",bins = 20)+
        geom_vline(xintercept = xi,col="red",linetype = "dashed")+theme_bw() + 
        theme(plot.background =    element_blank(),panel.grid.major = element_blank(),aspect.ratio = 0.7,
              panel.grid.minor = element_blank()) + xlab("$\\xi$") +ylab("Probability") + scale_x_continuous(limits = c(0, 5))+ggtitle("(a) Simulated Parameters") +theme(plot.title = element_text(size = 10),plot.title.position = "plot")



plot_grid(p18,p17,nrow = 1,rel_widths = c(1,1),align = "vh")

```

\begin{tcolorbox}[colback=cornsilk,colframe=aureolin]
Non-parametric, parametric and bayesian inference all demonstrate that the shape paramater $\xi$ is >0 and almost $85\%$ to $95\%$ probability that they are greater than $1$. This analysis confirms that pandemics are fat tailed phenomena. 
\end{tcolorbox}


```{r bayes1,echo=F,eval=F}
library(gridExtra)
library(rstan)
library(bayesplot)
library(loo)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores()-4)
options("scipen"=100, "digits"=4)

d <- data.frame(sort(dual))
colnames(d) <- "dual"


n <- dim(d)[1]
yt<- d$dst
ds<-list(ymin=200, N=n, Nu = 25, Nall = 72,H=7700000,L=200, y=d$dual, Nt=length(yt), yt=yt)
fit_gpd <- stan(file='r_stan.stan', data=ds, refresh=0,
                chains=4, seed=100)
params <- rstan::extract(fit_gpd)

```

```{r stan,echo=F,eval=F}
//functions for generalized pareto distribution and inverse gamma
//https://mc-stan.org/users/documentation/case-studies/gpareto_functions.html

functions {
  real gpareto_lpdf(vector y, real ymin, real k, real sigma) {
    // generalised Pareto log pdf
    int N = rows(y);
    real inv_k = inv(k);
    if (k<0 && max(y-ymin)/sigma > -inv_k)
      reject("k<0 and max(y-ymin)/sigma > -1/k; found k, sigma =", k, sigma);
    if (sigma<=0)
      reject("sigma<=0; found sigma =", sigma);
    if (fabs(k) > 1e-15)
      return -(1+inv_k)*sum(log1p((y-ymin) * (k/sigma))) -N*log(sigma);
    else
      return -sum(y-ymin)/sigma -N*log(sigma); // limit k->0
  }
  real gpareto_cdf(vector y, real ymin, real k, real sigma) {
    // generalised Pareto cdf
    real inv_k = inv(k);
    if (k<0 && max(y-ymin)/sigma > -inv_k)
      reject("k<0 and max(y-ymin)/sigma > -1/k; found k, sigma =", k, sigma);
    if (sigma<=0)
      reject("sigma<=0; found sigma =", sigma);
    if (fabs(k) > 1e-15)
      return exp(sum(log1m_exp((-inv_k)*(log1p((y-ymin) * (k/sigma))))));
    else
      return exp(sum(log1m_exp(-(y-ymin)/sigma))); // limit k->0
  }
  real gpareto_lcdf(vector y, real ymin, real k, real sigma) {
    // generalised Pareto log cdf
    real inv_k = inv(k);
    if (k<0 && max(y-ymin)/sigma > -inv_k)
      reject("k<0 and max(y-ymin)/sigma > -1/k; found k, sigma =", k, sigma);
    if (sigma<=0)
      reject("sigma<=0; found sigma =", sigma);
    if (fabs(k) > 1e-15)
      return sum(log1m_exp((-inv_k)*(log1p((y-ymin) * (k/sigma)))));
    else
      return sum(log1m_exp(-(y-ymin)/sigma)); // limit k->0
  }
  real gpareto_lccdf(vector y, real ymin, real k, real sigma) {
    // generalised Pareto log ccdf
    real inv_k = inv(k);
    if (k<0 && max(y-ymin)/sigma > -inv_k)
      reject("k<0 and max(y-ymin)/sigma > -1/k; found k, sigma =", k, sigma);
    if (sigma<=0)
      reject("sigma<=0; found sigma =", sigma);
    if (fabs(k) > 1e-15)
      return (-inv_k)*sum(log1p((y-ymin) * (k/sigma)));
    else
      return -sum(y-ymin)/sigma; // limit k->0
  }
  real gpareto_rng(real ymin, real k, real sigma) {
    // generalised Pareto rng
    if (sigma<=0)
      reject("sigma<=0; found sigma =", sigma);
    if (fabs(k) > 1e-15)
      return ymin + (uniform_rng(0,1)^-k -1) * sigma / k;
    else
      return ymin - sigma*log(uniform_rng(0,1)); // limit k->0
  }
  real gammainc(real a, real x){
    return gamma_q(a,x) * tgamma(a);}
}

//data vector

data {
  real ymin;
  int<lower=0> N;
  vector<lower=ymin>[N] y;
  int<lower=0> Nt;
  vector<lower=ymin>[Nt] yt;
}
transformed data {
  real ymax = max(y);
}
parameters {
  real<lower=0> sigma;
  real<lower=-sigma/(ymax-ymin)> k;
}
model {
  y ~ gpareto(ymin, k, sigma);
}
generated quantities {
  vector[N] log_lik;
  vector[N] yrep;
  vector[Nt] predccdf;
  
  
  for (n in 1:N) {
    log_lik[n] = gpareto_lpdf(rep_vector(y[n],1) | ymin, k, sigma);
    yrep[n] = gpareto_rng(ymin, k, sigma);
  }
  
  for (nt in 1:Nt)
    predccdf[nt] = exp(gpareto_lccdf(rep_vector(yt[nt],1) | ymin, k, sigma));
}

```


# Expected Value of Pandemic Fatalaities

We can estimate the Expected value due to dual transformation even though the shape parameter $\xi > 1$. Plugging in estimated values into equation 4 we obtain an estimate of 20.2 million casualties compared to a naive estimate of 13.96 million for data greater than 200,000. These values are summarized in Table \ref{table:tab_ey}. If we had used naive estimates we would have significantly underestimated the casualties.


\begin{table}[h!]
\caption{Table Comparing naive sample mean and expected value from GPD dual data. We observe that naive sample mean under estimates the casualties by 35$\%$ less than the GPD's expected value.}
\centering
\begin{tabular}{llll}
\hline
Estimate (in millions)        & Naive Sample & GPD (dual data) & Difference \\ \hline
$\mu$ (\textgreater{}200,000) & 13.96        & 20.2            & -35$\%$    \\ \hline
\end{tabular}
\label{table:tab_ey}
\end{table}

```{r,echo=F}
nu = 25
n = 72
H = 7700000
u = 200
xi = round(opt.par[2],2)
beta = round(opt.par[1],2)

sigma = beta*(nu/n)^xi
mu = u - ((beta/xi)*(1-((nu/n)^xi)))
L=200

Ey = ((H-L)*exp(sigma/(xi*H))*((sigma/(H*xi))^(1/xi))*expint::gammainc(1-(1/xi),(sigma/(H*xi))))+L

sample_mean = mean(dat[dat>=200])
```

# Conclusions and Implications

In this study we successfully replicated the work of @cirillo2020tail. We developed almost all the analysis from scratch and not relying on any canned functions from standard packages. 

Pandemics exhibit fat tail properties and therefore very hard to predict outcomes. This should be considered for risk management and decision making.


# References

<div id="refs"></div>

# Appendix

`R` code to reproduce models.

## Replicate Figure 1

```{r code1, ref.label='fig0', eval = FALSE}
```

## Replicate Figure 2

```{r code2, ref.label='fig1', eval = FALSE}
```

## Replicate Figure 3

```{r code3, ref.label='fig2', eval = FALSE}
```

## Replicate Figure 4

```{r code4, ref.label='fig3', eval = FALSE}
```


## Replicate Figure 5

```{r code5, ref.label='fig3a', eval = FALSE}
```

## Replicate Figure 8

```{r code6, ref.label='est', eval = FALSE}
```

## Replicate Figure 9

```{r code7, ref.label='nparm', eval = FALSE}
```

## Replicate Figure 10

```{r code8, ref.label='prof', eval = FALSE}
```

## Replicate Figure 11

```{r code9, ref.label='bayes', eval = FALSE}
```

```{r code10, ref.label='bayes1', eval = FALSE}
```

```{r code11, ref.label='stan', eval = FALSE}
```


```{r tab,message=FALSE,echo=F} 

import_data <- read.xlsx(file = "/mnt/home/user/stat508/final_project/data/long_fat_tail.xlsx",1)

colnames(import_data) <- c("Name","Start Year","End Year",
                            "Lower Est $\\times 10^3$", "Avg Est $\\newline \\times 10^3$","Upper Est $\\times 10^3$",
                              "Rescaled Avg Est $\\times 10^3$","Population $\\times 10^6$","Source")


import_data[,-9] %>%
  kable(booktabs = T, longtable = F,format = "latex", caption = "Table used in the data analysis. Please consult Cirillo and Taleb (2020) for references",linesep = "",escape = F) %>%   kable_styling(latex_options="scale_down", font_size = 11)
```













